---
title: "PreProcessing"
author: "MSProfileR"
output:
  html_document:
    code_folding: hide
    toc_float: yes
  pdf_document: default
params:
  spectraPath: ../MSProfileR-test/data/cullex_monique
  annotationFile: data/Mosquitoes/annotations_mosquitoes_ref.csv
  TIC: PQN
  SNR: '4'
  RData: a_output/a_preprocess.RData
  verbose: '0'
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# MODULE I : DATA LOADING
##############################################################################################################################
# Load Spectra, Retrieve Annotations

All spectra files are in the `r params$spectraPath` folder. All spectra are processed together.

```{r load}
library(MSProfileR)
library(MALDIquantForeign)
library(MALDIquant)
library(MALDIrppa)

## Load spectra 

spectra_raw <- import(params$spectraPath, verbose=FALSE)

length(spectra_raw)


### Creation of dataframe containing metaData of spectra -> The experimental informations

df.spectra <- create_df_spectra(spectra = spectra_raw)

library(knitr)
library(ggplot2)
``` 

# MODULE II : PREPROCESSING
################################################################################################################################

# spectra trimming

```{r spectra trimming} 
# search how to 7insert intervals (2-20 kda)
conformity_tests_results(spectra_raw)

spectra <- trim(spectra_raw)
conformity_tests_results(spectra)

# test trimming
spectra <- trim(spectra_raw, c(2500, 9000))
conformity_tests_results(spectra)

range(sapply(spectra, function(s) min(mass(s))))
range(sapply(spectra, function(s) max(mass(s))))

# back to full data
spectra <- trim(spectra_raw)

plot(spectra_raw[[47]], sub="")
plot(spectra[[47]], sub="")
```


# Conformity test

```{r conformity tests}
## Conformity tests
conformity_tab <- conformity_tests_results(spectra)
kable(conformity_tab)
```

# Spectra cleaning

```{r Negative intensity values are replaced by zeros}

# Processing spectra ----

spectra <- spectra_raw

if (params$verbose == "2") {
  plot_spectra_debug <- function() {
    plot(spectra[[1]])  # Full spectrum in original scale
    tmp <- spectra[[1]]
    tmp@mass <- sqrt(tmp@mass)
    # tmp@intensity <- log10((tmp@intensity))-1.5
    plot(tmp, type = "l")  # Part of spectrum in log10 scale
    plot(tmp, xlim = c(60, 80), type = "l")  # Part of spectrum in log10 scale
  }
}
if (params$verbose == "2") plot_spectra_debug()

## transform intensities
spectra <- get.transformed.spectra(spectra_raw) # spectra cleaning###############

if (params$verbose == "2") plot_spectra_debug()

gc()
```


# Quality control 

```{r, spectra screening (threshold)}

## Control Quality of Spectra

screenResults <- screenSpectra(spectra, estimator = "MAD", threshold = 0.9)

plot_screening.results(df.spectra = df.spectra, sc.results = screenResults)
```

# Spectra averaging

```{r, spectra averaging}

avgSpectra <- averageMassSpectra(spectra, labels=df.spectra$sample.name,
 method="mean")
length(avgSpectra)
# View(avgSpectra)

```



## Peak detection

The spectra are continuous curves representing peaks. The center and the intensity of each peak need to be determined to simplify the information. For each spectrum this detection results in a peak list.

```{r, peak detection with number (without graphic)}

peaks <- get.peaks(avgSpectra, as.numeric(params$SNR))

plot(avgSpectra[[16]], xlim=c(4000, 5000), ylim=c(0, 0.002))
plot(peaks[[16]], col="red", pch=4)

length(peaks)
sapply(peaks, length)
hist(sapply(peaks, length))
```

# Spectra re-aligning

Because spectra are not perfectly aligned (calibration dispersion?), the most intense peaks in common through the data set are selected and used to realign the mass and the spectra.

```{r, spectra aligning}

# Got peaks for each spectrum
# Search for reference peaks across all spectra
p_ref <- referencePeaks(peaks, method = "strict", minFrequency = 0.3, tolerance = 2500e-6)
p_ref

# Define functions to realign each spectra to the reference peaks
w_funs <- determineWarpingFunctions(peaks, p_ref, tolerance = 2500e-6, method = "linear", allowNoMatches = TRUE)
# initially tol=4000ppm
# QC
plot(attr(w_funs, "nmatch"), type = "p", las = 1, ylab = "# matched peaks", xlab = "spectrum id", main = "Count of peaks matched to reference", ylim = c(0, length(p_ref)*1.1), cex = 0.8)
abline(h = length(p_ref) * c(1, 0.9, 0.8), lw = 2, col = c("#44EE4488", "#EE884488", "#EE444488"))
legend("bottomright", sprintf("%3d%%", c(1, 0.9, 0.8)*100), lwd = 3, col = c("#44EE4488", "#EE884488", "#EE444488"), xjust = 1, title = "% match")

# Align peak lists
p_aligned <- warpMassPeaks(peaks, w_funs, emptyNoMatches = TRUE)

```


```{r, spectra binnig}
p_aligned_bin <- binPeaks(p_aligned, tolerance=0.002)
p_aligned_bin_filt <- filterPeaks(p_aligned_bin, minFrequency=0) ###########" freq 0 à corriger

featureMatrix <- intensityMatrix(p_aligned_bin_filt, avgSpectra)
head(featureMatrix[, 1:6])
dim(featureMatrix)

image(sqrt(t(featureMatrix)), col = grey.colors(64), xaxt = "n", yaxt = "n", xlab = "m/z", ylab = "samples", mgp = c(0, 0, 0))
m.mass <- as.numeric(colnames(featureMatrix))

```





# Spectra clustring (heatmap and dendrogram)

Usually a heatmap is built to get an overview of the detected and aligned peaks. Peaks frequently present in spectra are kept, others are filtered out.

```{r, spectra clustring}
# Create a heatmap view of peak lists ----

if (FALSE) {
# Peaks that enter the same bin are matched and assigned an average mass
# try relaxed? p_binned <- binPeaks(p_aligned, method = "relaxed", tolerance = 1800e-6)
# p_binned <- binPeaks(p_aligned, method = "strict", tolerance = 1800e-6)
p_binned <- binPeaks(p_aligned, method = "strict", tolerance = 1000e-6)
# Lionel looks for tol = 1000ppm

# Frequent peaks are kept
p_filtered <- filterPeaks(p_binned, minFrequency = 0.5)

# A matrix of intensity can be extracted

s_aligned <- warpMassSpectra(spectra, w_funs, emptyNoMatches = TRUE)
p_matrix <- intensityMatrix(p_filtered, s_aligned)

dim(p_matrix)
kable(p_matrix[]) ############### 1:20, 1:7 inside the quote
round(prop.table(table(is.na(p_matrix))) * 100, 2)

# The heatmap of common peaks across spectra
# Each row is a sample. Each column is a peak. The intensity is encoded by the grey level, black meaning very intense. White corresponds to an undetected peak.
image(sqrt(t(p_matrix)), col = grey.colors(64), xaxt = "n", yaxt = "n", xlab = "m/z", ylab = "samples", mgp = c(0, 0, 0))
m.mass <- as.numeric(colnames(p_matrix))
# Each peak mass of the columns
m.mass



distanceMatrix <- dist(featureMatrix, method = "euclidean")
hClust_row <- hclust(distanceMatrix, method = "complete")

plot(hClust_row, hang = -1)



distanceMatrix <- dist(t(featureMatrix), method = "euclidean")
hClust_col <- hclust(distanceMatrix, method = "complete")

image(sqrt(t(featureMatrix[hClust_row$order, hClust_col$order])), col = grey.colors(64), xaxt = "n", yaxt = "n", xlab = "m/z", ylab = "samples", mgp = c(0, 0, 0))
image(sqrt(t(featureMatrix[hClust_row$order, ])), col = grey.colors(64), xaxt = "n", yaxt = "n", xlab = "m/z", ylab = "samples", mgp = c(0, 0, 0))

}


# install.packages("pheatmap")

avgSpectra_sample = data.frame(
  sample.name = sapply(avgSpectra, function(s) s@metaData$sampleName)
)

df.tmp = unique(df.spectra[, c("folder.name", "sample.name", "body.part")])

df.avgSpectra = merge(avgSpectra_sample, df.tmp, sort = FALSE)


rownames(featureMatrix) <- df.avgSpectra$sample.name
rownames(df.avgSpectra) <- df.avgSpectra$sample.name

df.avgSpectra$body.part = substr(df.avgSpectra$sample.name, 1, 2)

dim(featureMatrix)

pheatmap::pheatmap(
  sqrt(featureMatrix), cluster_rows = TRUE, cluster_cols = FALSE, show_rownames = FALSE, show_colnames = FALSE, annotation_row = df.avgSpectra[, "body.part", drop = FALSE]
)


```


## Snapshot

```{r echo=TRUE, eval=FALSE}
# save(merge_final, annot_final, spectra, s, p, p_ref, # annot, clinics, 
#      p_aligned, p_binned, p_filtered, p_matrix,
#      s_aligned, new_mz_nb, mz_new, mat_aligned, mat_unaligned, file = params$RData)
output_dir <- dirname(params$RData)
if (!dir.exists(output_dir)) dir.create(output_dir)
save(
  params,
  spectra, p_ref, 
  p_aligned, p_binned, p_filtered, p_matrix,
  s_aligned, file = params$RData
)
```
